# Exercise 1
<!-- '- Set up Rstudio Project -->
<!-- - Import and clean data -->
<!-- - Explore Dataset: remove outliers, find sampling interval, make subsets and find overlapping areas of individuals) -->
<!-- - Make my first simple map -->

Exercise 1 covers the necessary steps for getting ready in `R` and some basic concepts for setting up a well-structured `R` project. The lesson introduces how additional packages that provide useful functions for data science are made available and how spatial data is handled. The exercise concludes with the creation of your first map featuring movement data.


## Leaning outcomes
- You learn how to structure an `R` project.
- You can read movement data from a .csv-file into a `data.frame`
- You can convert spatial point data from a `data.frame` to a spatial object `sf`
- You can perform basic spatial operations on spatial objects in `R`
- You can produce simple maps of your spatial data using `ggplot2`


## Prerequisites
Readings Skills from "R for Data Science" [@wickham2017]:

- RS1.1 Preface (16p, ix-xxiv)
- RS1.2 Chap2 Workflow basics (3p, 37-39)
- RS1.3 Chap4 Workflow scripts (3p, 77-79)
- RS1.4 Chap6 workflow projects (6p, 111-116)
- RS1.5 Chap8 Data Import with `readr` (21p)
- RS1.6 Chap13 Date and Times with `lubridate` (18p, 237-256)


## Preperation

If you haven't already, install the packages `tidyverse`, and `devtools` (using `install.packages()`). Use `devtools` to install the newest version of `ggplot`[^2] from github (see code below). Additionally, install the packages `sf`[^1], `raster` and `ggspatial`. **Restart your `R` session after installing all these packages.**

[^1]: If you are working on RStudio Server, you will need to install `udunits2` with some additional parameters (see code example). 
[^2]: We will need the newest version of `ggplot2` to plot spatial `sf` objects.

```{r, echo = T, include=T,eval = F}
install.packages("tidyverse")
install.packages("devtools")
devtools::install_github("tidyverse/ggplot2")

# run the following line only if you are working on RStudio Server
install.packages('udunits2', type = "source", repo = "cran.rstudio.com", configure.args="--with-udunits2-include=/usr/include/udunits2")

install.packages("sf")



install.packages("raster")
install.packages("ggspatial")
```


## Tasks and inputs

### Task 1: Initialize project

Create a new *RStudio Project*. As recommended in @wickham2017, remove the option "*Restore .RData into workspace at startup*" and set the option "*save workspace to .RData on exit*" to "*Never*".

Create a new R-File and divide it into the sections necessary in a classical Data Science Workflow. We recommend following sections:

- Loading environment / libraries
- Data import
- Data cleansing
- Data analysis and visualization

"Sections" can be created within RStudio by adding Comments (`#`) with at least 4 trailing dashes (`-`), equal signs (`=`), or pound signs (`#`).  Sections allow code folding (try clicking on the small triangle next to the line number) and facilitate navigation (try the shortcut: `Shift`+`Alt`+`J`).

Load the package `tidyverse` in the appropriate section. 

```{r}

# Loading enironment / libraries ####
library(tidyverse)



# Data import ####

# Data clensing ####

# Data analysis and visualization ####

```


### Task 2: Import data

In section "data import", import the file `wildschwein_BE.csv`. 

Note: 

- If your are using [a graphical tool](https://support.rstudio.com/hc/en-us/articles/218611977-Importing-Data-with-RStudio) to import your code, make sure you save the corresponding code in your R Script. This is important in regard to the reproducibility of your script and will ensure that your workflow is documented without gaps. We'd rather recommend to move away from using graphical tools and focus on using code. 
- We recommend using one of the `tidyverse` functions (`read_*`) to import your data. These functions are less error prone than the base `R` functions (`read.*`). Specifically for the wild boar data, we recommend `read_delim()`.
- If you use `read_delim()` and receive warnings during import, have a look at these warnings by using the function `problems()`. Resolve these problems until import runs without warnings.
- Assign correct data types as necessary and make sure the time zone is set correctly for the date/time column.
- For everyone working on the RStudio Server: You will first need to upload this data to the server using the "*upload*"-button in the "*Files*" tab.


```{r}


## Task 2 ####################


# Data import ####
wildschwein_BE <- read_delim("../CMA_FS2018_Filestorage/wildschwein_BE.csv",",")


# Check Timezone
attr(wildschwein_BE$DatetimeUTC,"tzone") # or
wildschwein_BE$DatetimeUTC[1]
```



### Task 3: Explore Data
We will use a range of different plotting techniques in this course. Several plotting techniques have emerged in recent years, each with their specific strengths and weaknesses. While `base::plot() `is quick and simple, it not very scalable with growing complexity. `ggplot` offers solutions for most use cases and has a elegant syntax that is easy to get accustomed to. We will get to know other techniques later in the course.

Get an overview of your data by creating a first "map-like" plot of your data producing a simple scatter plot with `ggplot`. Do you spot outliers? If so, get rid of the outliers and plot your data again without outliers. 

Setting up a `ggplot` with our data is done using the command `ggplot(wildschwein_BE, aes(Long, Lat, colour = TierID))`. Creating a map is done via the basic scatter plot command `geom_point()`. Use `coord_map()` to get a reasonable aspect ratio of `Lat` and `Long`.  Assigning every individual its own colour is done using the `ggplot` argument `colour =`.

Save your code in the appropriate section.

```{r, echo = F, include=T, eval = T, }

## Task 3 ####################
ggplot(wildschwein_BE, aes(Long,Lat, colour = TierID)) +
  geom_point() +
  coord_map() +
  theme(legend.position = "none")
```





### Input: Handling spatial data

Until now, we've handled spatial data within data frames. This works well for many tasks, but sometimes we need special *spatial* classes to handle our trajectories. We will get to now such cases in our next tasks, but first we need to convert our `data.frame` into a spatial object.
Some of you might be familiar with the `sp` package with the classes `SpatialPoints`, `SpatialPointsDataFrame` and so on. Just recently the new and exiting package `sf`, was released on CRAN. `sf` has some huge advantages over sp:

- simple features are essentially data frames, which mean they interface with the `tidyverse`  
- comply with the common Open Geospatial Consortium (OGC) standards (ISO 19125-1:2004) and interface with other important spatial tools such as GDAL, PostGIS, GeoJSON and so fourth
- are being rapidly implemented in visualisation tools such as `ggplot2`, `plotly` and `tmap`


We will largely rely on `sf`when working with vector data in `R`. In order to transform our `data.frame` into an sf object, we need to use the function `st_as_sf()` while specifying the columns storing the coordinates and the coordinate reference system[^10].

[^10]: At this point, we assume you know what a Coordinate Reference Systems is. Check out [this link](https://earthdatascience.org/courses/earth-analytics/spatial-data-r/intro-to-coordinate-reference-systems/) if this is not the case. 

```{r}
## Input: Handling spatial data ####################
```



```{r, echo = T, include = T, eval = T}

library(sf)

wildschwein_BE_sf <- st_as_sf(wildschwein_BE, coords = c("Long", "Lat"), crs = 4326)

```

Notice how `st_as_sf` takes the EPSG code for the `crs =` argument. This is so much easier and more elegant than using `PROJ.4` or `WKT`. You can find a lot of useful information on Coordinate Reference Systems (including EPSG Codes , etc.) under [spatialreference.org](http://spatialreference.org/ref/epsg/2056/) or http://epsg.io.  

Let's compare our original `data.frame` with this new `sf` object:

```{r,  echo = T, include = T, eval = T, collapse=F}
wildschwein_BE

wildschwein_BE_sf
```

As you can see, `st_as_sf()` has added some metadata to our dataframe (`geometry type`, `dimension`, `bbox`, `epsg` and `proj4string`) and replaced the columns `Lat` and `Long` with a column named `geometry`. Other than that, the new `sf` object is very similar to our original dataframe. In fact, `sf` objects *are* essentially `dataframes`, just ask `R`:

```{r,  echo = T, include = T, eval = T}
is.data.frame(wildschwein_BE_sf)
```

All operations we know from handling `data.frames` can be used on the `sf` object. Try some out!
```{r,  echo = T, include = T, eval = F}
# subset rows
wildschwein_BE_sf[1:10,]
wildschwein_BE_sf[wildschwein_BE_sf$TierName == "Ueli",]

# subset colums
wildschwein_BE_sf[,2:3]
```

Instead of keeping the same data twice (once as a `data.frame`, and once as an `sf` object), we will overwrite the `data.frame` and continue working with the `sf` object from now on. This saves some memory space in `R` and avoids confusion. It is however, unideal that the `st_as_sf()` operation *replaced* the `Lat` and `Long` columns with the `geometry` column. It would be helpful to retain the `Lat`/`Long` columns and have the `geometry` column *in addition*. We can enforce this by using the argument `remove = FALSE`.

```{r,  echo = T, include = T, eval = T}
wildschwein_BE = st_as_sf(wildschwein_BE, coords = c("Long", "Lat"), crs = 4326,remove = FALSE)

wildschwein_BE # note how the Lat/Long information is stored twice

rm(wildschwein_BE_sf) # we can remove this sf object, since it just eats up our memory

```


### Task 4: Project data from WGS84

So what can we do with our new `sf` object that we couldn't before? One example is projecting the WGS84 (`Lat`/`Long`) coordinates into the new Swiss CRS `CH1903+ LV95`[^50]. Do this by using the function `st_transform`. By the way, do you notice a pattern here? The package `sf` names most functions for spatial operations with the prefix `st_*`.

[^50]: As we've mentioned in the first Input, you can look up the EPSG codes under [spatialreference.org](http://spatialreference.org/ref/epsg/2056/) or http://epsg.io. For information specific to switzerland switzerland, check the [swisstopo website](https://www.swisstopo.admin.ch/en/knowledge-facts/surveying-geodesy/reference-systems.html) 

```{r}

## Task 4 ####################

wildschwein_BE <- st_transform(wildschwein_BE, 2056)


```

Here's the resulting `sf` object from the operation:
```{r, echo = F, include=T,eval = T}
wildschwein_BE
```



### Input: Calculate Convex Hull

Transforming from one Coordinate Reference System to another was one operation where we needed an object with a spatial nature. In this way, we were able to use an off the shelf function to project the coordinates from one CRS to another. In our next example, we again rely on a spatial function: We want to calculate a [convex hull](https://en.wikipedia.org/wiki/Convex_hull) per Wild boar. And guess what the function for calculating a convex hull is called in `sf`? If you fuessed `st_convex_hull()`, you were right!

Before we can ferform `st_convex_hull()`, we need to add a grouping variable to our `sf` object, so that the convex hull is calculated *per animal*. Note:

- Grouping (via `group_by()`) has no effect on our original data frame
- Effects of `group_by()` only take effect when manipulations are performed on the data.frame via `summarise()` or `mutate()`
- If you know `summarise()` and `mutate()` operations from `dplyr`; `sf` understands this language which facilitates handling spatial data 

```{r}
## Input: Convex Hull ####################
```


```{r, echo = T,include = T,eval = T}
wildschwein_BE <- group_by(wildschwein_BE,TierID)
```

You can see the grouping variable in the metadata of our `sf` object:

```{r, echo = T,include = T,eval = T}
wildschwein_BE
```

In order for `sf_convex_hull()` to respect the grouping variable, we need to warp the `sf` object within `summarise()`

```{r, echo = T,include = T,eval = T}
mcp <- st_convex_hull(summarise(wildschwein_BE))

```


### Task 5: Ploting spatial objects

Using base plot to visualize `sf` objects is easy enough, just try the following code. 

```{r}
## Task 5 ####################
```


```{r, echo = T,include = T,eval = T}
plot(mcp)
```

But since we use `ggplot` extensively, try and plot the object `mcp` with `ggplot`. Hint: Since you installed the newest version from github, you now use the function `geom_sf()` to add an `sf` object.

```{r, echo = F,include = T,eval = T}
ggplot(mcp,aes(fill = TierID)) +
  geom_sf(alpha = 0.4)
```

Note: 

- `ggplot` refuses to use our specified CRS, so we need to force this by specifying `coord_sf()`
- Oddly, it is not sufficient to specify the `crs = ` wit hin `coord_sf()`, we need to pass our EPSG code to `datum =`. 
- You can change the plot style by making adjustments within `theme()` I recommend the following adjustments:


```{r, echo = T,include = T,eval = F}
theme(
  legend.position = "none",
  panel.grid.major = element_line(colour = "transparent"),
  panel.background = element_rect(fill = "transparent"),
  axis.title = element_blank(),
  axis.text = element_blank(),
  axis.ticks = element_blank()
  )
```


```{r, echo = F,include = T,eval = T}
ggplot(mcp,aes(fill = TierID)) +
  geom_sf(alpha = 0.4) +
  coord_sf(datum = 2056) +
  theme(
    legend.position = "none",
    panel.grid.major = element_line(colour = "transparent"),
    panel.background = element_rect(fill = "transparent"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
    )
```


### Task 6: Adding a background map


In order to add a background map to our plot, we will need to load two additional libraries: `raster` to load raster files and `ggspatial` to plot raster files in `ggplot`. Load these libraries now, and import the file `pk100_BE_2056.tif` (available on moodle) using the function `brick()`. Add the newly created `RasterBrick` object to ggplot using the function `geom_spraster_rgb()`.
 

```{r, echo = F, include = T, eval = T}

## Task 6 ####################


library(raster)
library(ggspatial)

pk100_BE <- brick("../CMA_FS2018_Filestorage/pk100_BE_2056.tif")

ggplot(mcp,aes(fill = TierID)) +
  geom_spraster_rgb(pk100_BE, interpolate = TRUE) +
  geom_sf(alpha = 0.4) +
  coord_sf(datum = 2056) +
  theme(
    legend.position = "none",
    panel.grid.major = element_line(colour = "transparent"),
    panel.background = element_rect(fill = "transparent"),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
    )

```


<!-- ## Solutions (RCode) -->

<!-- ```{r code=readLines('11_Week1/RFiles/W01_01_Exercise.R'), results='asis', echo = T, include=T, eval=F} -->
<!-- ``` -->